pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-02-01 bmc_article.tex

\title{A framework for pipeline benchmarking and its application to single-cell RNAseq clustering}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit extensible benchmarks (\url{https://github.com/plger/pipeComp}).

@@

While a number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}), they need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate the effect of the choice of a tool on downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}). It is then likely that processing algorithms also interact in a positive or negative fashion and it is then critical to evaluate not only the single effect of a method but also its interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and developed new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

Another motivation for the present framework is that clustering is relatively fragile to variations in the pipeline. We therefore also wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We therefore proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the pipeComp framework and its application to a scRNAseq clustering pipeline.}}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C. We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in feature selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives. Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next investigated the methods of dimensionality reduction. Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization. Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.

@@

In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology. Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking. Differences between such protocols can have a very large impact on clustering \citep{MereuCellAtlas2019}. Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies. In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \citep{viethSystematic2019}. Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019, TsuyuzakiPCA2020}. We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.

Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}). In contrast to Tian et al. \citep{tianMixology2018}, who used on RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability. Given the limited availability of such datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration. For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}. In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference. Several informative benchmarks have already been performed on some of these topics \citealp{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}. Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here. We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.

@@

\section*{Competing interests}
  The authors declare that they have no competing interests.

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & source & protocol & description \\ 
  \hline
Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}

>>>>

\title{pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, permit extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

@@

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}). It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline. We therefore favour monitoring several complementary metrics across multiple steps of the process.

A clustering output is relatively fragile to variations in the pipeline. We therefore wanted to capture whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline, or rather specific to a set of other pipeline parameters. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:figure2}B).

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept of number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies, and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

@@

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions ($\geq$3.0) rely on a different measure of variance, again standardized. While adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in feature selection is to use those with the highest mean expression. Recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three of them. Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

Since the various PCA approaches and implementations were recently benchmarked in a similar context \citep{TsuyuzakiPCA2020}, we focused on widely used approaches which had not yet been compared: \textit{Seurat}'s PCA, \textit{scran}'s \texttt{denoisePCA}, and GLM-PCA \citep{townesGlmpca2019}, and combined them (when relevant) with \textit{sctransform} normalization. Given that \textit{Seurat}'s default PCA weights the cell embeddings by the variance of each component, we also evaluated the impact of this weighting with each method.

@@

In this study, we evaluated tools commonly used for the processing of scRNAseq with a focus on droplet-based datasets, namely from the 10x technology. Although this platform has been used in almost half of the scRNAseq studies in 2019 \citep{SvenssonDB2019}, other popular technologies such as Drop-seq, InDrops or Smart-seq2/3 were not represented in the present benchmarking. Differences between such protocols, even among droplet-based technologies, can have a very large impact on cell capture efficiency, cell numbers and clustering \citealp{MereuCellAtlas2019, ZhangDroplet2019, SalomonDroplet2019}. Although most top-ranking methods in our comparison performed well on both Smart-seq and 10x datasets that we tested, future benchmarking efforts should strive to include less represented technologies. In addition, we did not compare any of the alignment and/or quantification methods used to obtain the count matrix, which was for instance discussed in \citep{viethSystematic2019}. Some steps, such as the implementation of the PCA, were also not explored in detail here as they have already been the object of recent and thorough study elsewhere \citep{SunDimRed2019, TsuyuzakiPCA2020}. We also considered only methods relying on Euclidean distance, while correlation was recently reported to be superior\citep{kim_impact_2019} and would require further investigation.

Here, we chose to concentrate on what could be learned from datasets with known cell labels (as opposed to labels inferred from the data, as in \citep{MereuCellAtlas2019}). In contrast to Tian et al. \citep{tianMixology2018}, who used RNA mixtures of known proportions, we chose to rely chiefly on real cells and their representative form of variability. Given the limited availability of such well-described datasets however, several aspects of single-cell analysis could not be compared, such as batch effect correction or multi-dataset integration. For these aspects of scRNAseq processing that are critical in some experimental designs, we refer the reader to previous evaluations \citealp{TranBatch2020, StuartIntegration2019}. In addition, a focus on the identification of the subpopulations might fail to reveal methods that are instead best performing for tasks other than clustering, such as differential expression analysis or trajectory inference. Several informative benchmarks have already been performed on some of these topics \citealp{CrowellMuscat2019, DalMolinDE2017, JaakkolaDE2017, SaelensTraject2019, SonesonDE2018, WangDE2019}. Yet, such evaluations could benefit from considering methods not in isolation, but as parts of a connected workflow, as we have done here. We believe that the \textit{pipeComp} framework is modular and flexible enough to integrate new steps in the pipeline, as shown by an example with imputation/denoising methods.

We developed \ŧexttt{pipeComp} to address the need of a framework that can evaluate multiple tools simultaneously. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. A key additional feature of \textit{pipeComp} is its ability to perform evaluation in parallel to the pipeline and offering on-the-fly evaluation at any step. This avoid the need to store potentially large intermediate data for all possible permutations and thus allowing a combinatorial benchmark. In addition, the fact that benchmark functions are stored in the \texttt{PipelineDefinition} makes the benchmark more smoothly portable, enabling it transportable easy to apply comparable contains all information regarding the benchmark, including

@@

\section*{Competing interests}
As developers of the \textit{scDblFinder} package, the authors have an interest in it performing well. The authors declare no further competing interests.

@@

\begin{table}[h!]
\caption{Overview of the benchmark datasets.}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & source & protocol & description \\ 
  \hline
Koh & GSE85066 & SMARTer & FACS purified H7 hESC in different differention stages \\ 
  Kumar & GSE60749 & SMARTer & Mouse ESC cultured in different conditions \\ 
  Zhengmix4eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & \citep{duoClustering2018} & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & \cite{tianMixology2018} & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & \cite{tianMixology2018} & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & - & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & - & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}
